---
title: "Dimensionality Reduction with UMAP"
subtitle: "Complementing Categorical Encoding with uwotlite"
author: "Based on techniques from Tidy Modeling with R"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Dimensionality Reduction with UMAP}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)

# Check for required packages (including uwotlite itself for vignette building)
required_pkgs <- c("uwotlite", "modeldata", "ggplot2", "dplyr", "scales")
have_pkgs <- vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
if (!all(have_pkgs)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

This vignette demonstrates UMAP (Uniform Manifold Approximation and Projection) dimensionality reduction using `uwotlite`, as a complement to the categorical encoding techniques shown in [Chapter 17 of Tidy Modeling with R](https://www.tmwr.org/categorical).

## Introduction

While Chapter 17 of TMWR focuses on encoding categorical variables into numeric representations, another common preprocessing task is *dimensionality reduction*—transforming many features into fewer features while preserving important structure.

UMAP is a nonlinear dimensionality reduction technique that:

- Preserves both local and global structure in data
- Scales well to large datasets
- Can be used for visualization (2D) or feature creation (higher dimensions)
- Supports supervised reduction using outcome information

The `uwotlite` package provides an MIT-licensed UMAP implementation.

## Setup

```{r}
#| label: libraries
library(uwotlite)
library(dplyr)
library(ggplot2)
library(modeldata)

# Load the Ames housing data
data(ames)

# Create train/test split
set.seed(502)
n <- nrow(ames)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
ames_train <- ames[train_idx, ]
ames_test <- ames[-train_idx, ]
```

## Preparing Numeric Features

After encoding categorical variables (using effect encodings as shown in the `embedmit` vignette), you often have many numeric features. Let's prepare a numeric dataset from the Ames data:

```{r}
#| label: prepare-data
# Select numeric features
ames_numeric <- ames_train %>%
  select(
    Lot_Area, Gr_Liv_Area, Year_Built, Year_Remod_Add,
    Total_Bsmt_SF, First_Flr_SF, Second_Flr_SF,
    Full_Bath, Half_Bath, Bedroom_AbvGr, TotRms_AbvGrd,
    Fireplaces, Garage_Cars, Garage_Area,
    Wood_Deck_SF, Open_Porch_SF
  ) %>%
  # Handle missing values
  mutate(across(everything(), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  # Scale the features
  mutate(across(everything(), scale))

# Convert to matrix for uwotlite
ames_matrix <- as.matrix(ames_numeric)

dim(ames_matrix)
```

We have `r nrow(ames_matrix)` observations and `r ncol(ames_matrix)` numeric features.

## Basic UMAP

The `umap()` function performs dimensionality reduction:

```{r}
#| label: umap-basic
set.seed(123)
ames_umap <- umap(
  ames_matrix,
  n_neighbors = 15,
  n_components = 2,
  metric = "euclidean",
  min_dist = 0.1,
  verbose = FALSE
)

# Combine with price for visualization
umap_df <- data.frame(
  UMAP1 = ames_umap[, 1],
  UMAP2 = ames_umap[, 2],
  Sale_Price = ames_train$Sale_Price,
  Neighborhood = ames_train$Neighborhood
)

head(umap_df)
```

## Visualizing the UMAP Embedding

### Colored by Sale Price

```{r}
#| label: umap-viz-price
#| fig-cap: "UMAP projection of Ames housing features colored by sale price. Similar houses (in terms of the 16 numeric features) are positioned near each other."
ggplot(umap_df, aes(UMAP1, UMAP2, color = Sale_Price)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_viridis_c(labels = scales::dollar) +
  labs(
    title = "UMAP Projection of Ames Housing Features",
    subtitle = "Colored by Sale Price",
    color = "Sale Price"
  ) +
  theme_minimal() +
  coord_fixed()
```

The UMAP projection reveals structure in the data—houses with similar characteristics cluster together, and there's a gradient of sale prices across the embedding.

### Colored by Neighborhood

```{r}
#| label: umap-viz-neighborhood
#| fig-cap: "UMAP projection colored by neighborhood. Some neighborhoods form distinct clusters while others overlap."
# Show top 8 neighborhoods for clarity
top_neighborhoods <- ames_train %>%
  count(Neighborhood, sort = TRUE) %>%
  head(8) %>%
  pull(Neighborhood)

umap_df %>%
  filter(Neighborhood %in% top_neighborhoods) %>%
  ggplot(aes(UMAP1, UMAP2, color = Neighborhood)) +
  geom_point(alpha = 0.6, size = 1.5) +
  labs(
    title = "UMAP Projection by Neighborhood",
    subtitle = "Top 8 neighborhoods shown"
  ) +
  theme_minimal() +
  coord_fixed()
```

## Key UMAP Parameters

Understanding the main parameters helps tune UMAP for your use case:

| Parameter | Description | Effect |
|-----------|-------------|--------|
| `n_neighbors` | Size of local neighborhood | Higher = more global structure, lower = more local detail |
| `min_dist` | Minimum distance between points | Higher = more spread out, lower = tighter clusters |
| `n_components` | Output dimensions | 2 for visualization, higher for feature creation |
| `metric` | Distance measure | "euclidean", "cosine", "manhattan", etc. |

### Effect of n_neighbors

```{r}
#| label: param-comparison
#| fig-cap: "Effect of n_neighbors on UMAP embedding. Lower values emphasize local structure, higher values capture more global patterns."
#| fig-height: 8
set.seed(123)

# Compare different n_neighbors values
neighbor_values <- c(5, 15, 50)
umap_results <- lapply(neighbor_values, function(nn) {
  result <- umap(ames_matrix, n_neighbors = nn, n_components = 2,
                 min_dist = 0.1, verbose = FALSE)
  data.frame(
    UMAP1 = result[, 1],
    UMAP2 = result[, 2],
    Sale_Price = ames_train$Sale_Price,
    n_neighbors = paste("n_neighbors =", nn)
  )
})

bind_rows(umap_results) %>%
  ggplot(aes(UMAP1, UMAP2, color = Sale_Price)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_color_viridis_c() +
  facet_wrap(~n_neighbors, scales = "free") +
  labs(title = "Effect of n_neighbors on UMAP Structure") +
  theme_minimal()
```

## Transforming New Data

A key feature of `uwotlite` is the ability to transform new data using a fitted UMAP model. This is essential for prediction workflows.

```{r}
#| label: umap-transform
# Fit UMAP with ret_model = TRUE to enable transformation
set.seed(123)
ames_umap_model <- umap(
  ames_matrix,
  n_neighbors = 15,
  n_components = 2,
  min_dist = 0.1,
  ret_model = TRUE,
  verbose = FALSE
)

# Prepare test data the same way
ames_test_numeric <- ames_test %>%
  select(
    Lot_Area, Gr_Liv_Area, Year_Built, Year_Remod_Add,
    Total_Bsmt_SF, First_Flr_SF, Second_Flr_SF,
    Full_Bath, Half_Bath, Bedroom_AbvGr, TotRms_AbvGrd,
    Fireplaces, Garage_Cars, Garage_Area,
    Wood_Deck_SF, Open_Porch_SF
  ) %>%
  mutate(across(everything(), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(everything(), scale))

ames_test_matrix <- as.matrix(ames_test_numeric)

# Transform test data
test_umap <- umap_transform(ames_test_matrix, ames_umap_model)

head(test_umap)
```

### Visualizing Train and Test Together

```{r}
#| label: train-test-viz
#| fig-cap: "UMAP projection showing both training and test data. Test data is transformed using the model fitted on training data."
train_df <- data.frame(
  UMAP1 = ames_umap_model$embedding[, 1],
  UMAP2 = ames_umap_model$embedding[, 2],
  Set = "Train"
)

test_df <- data.frame(
  UMAP1 = test_umap[, 1],
  UMAP2 = test_umap[, 2],
  Set = "Test"
)

bind_rows(train_df, test_df) %>%
  ggplot(aes(UMAP1, UMAP2, color = Set)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(
    title = "UMAP: Training vs Test Set Projection",
    subtitle = "Test data transformed using fitted model"
  ) +
  theme_minimal() +
  coord_fixed()
```

## Using UMAP Components as Features

UMAP components can serve as input features for downstream models:
```{r}
#| label: umap-features
# Create UMAP features for modeling
train_features <- data.frame(
  UMAP1 = ames_umap_model$embedding[, 1],
  UMAP2 = ames_umap_model$embedding[, 2],
  Sale_Price = ames_train$Sale_Price
)

# Simple linear model using UMAP features
umap_lm <- lm(Sale_Price ~ UMAP1 + UMAP2, data = train_features)
summary(umap_lm)
```

The R-squared tells us how much of the sale price variance is captured by just two UMAP dimensions.

## Supervised UMAP

`uwotlite` supports *supervised* UMAP, which uses label information to guide the embedding. This can improve the embedding for prediction tasks:

```{r}
#| label: supervised-umap
#| fig-cap: "Supervised UMAP uses the Sale Price outcome to guide the projection, potentially creating better separation for prediction tasks."
set.seed(123)
ames_supervised <- umap(
  ames_matrix,
  n_neighbors = 15,
  n_components = 2,
  min_dist = 0.1,
  y = ames_train$Sale_Price,
  target_weight = 0.5,
  verbose = FALSE
)

supervised_df <- data.frame(
  UMAP1 = ames_supervised[, 1],
  UMAP2 = ames_supervised[, 2],
  Sale_Price = ames_train$Sale_Price
)

ggplot(supervised_df, aes(UMAP1, UMAP2, color = Sale_Price)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_viridis_c(labels = scales::dollar) +
  labs(
    title = "Supervised UMAP Projection",
    subtitle = "Embedding guided by Sale Price"
  ) +
  theme_minimal() +
  coord_fixed()
```

## Integration with embedmit

For a complete workflow combining categorical encoding and UMAP, use `embedmit` with `step_umap()`:

```{r}
#| label: embedmit-integration
#| eval: false
library(embedmit)
library(recipes)

# Complete pipeline: encode categoricals, then reduce dimensions
ames_recipe <- recipe(Sale_Price ~ ., data = ames_train) %>%
  # Encode high-cardinality categorical
  step_lencode_mixed(Neighborhood, outcome = vars(Sale_Price)) %>%
  # Create dummy variables for remaining categoricals
  step_dummy(all_nominal_predictors()) %>%
  # Reduce to UMAP components
  step_umap(all_numeric_predictors(), num_comp = 5, neighbors = 15)
```

This pipeline:

1. Encodes `Neighborhood` using mixed effects (see the embedmit vignette)
2. Creates dummy variables for other categorical predictors
3. Reduces all numeric features to 5 UMAP components

## Summary

UMAP is a powerful dimensionality reduction technique that complements categorical encoding:

| Use Case | Technique |
|----------|-----------|
| Encode high-cardinality categoricals | Effect encodings (`step_lencode_*`) |
| Visualize high-dimensional data | UMAP with `n_components = 2` |
| Create features from many predictors | UMAP with `n_components > 2` |
| Improve embeddings for prediction | Supervised UMAP with `y` parameter |
| Apply to new data | `umap_transform()` with `ret_model = TRUE` |

The `uwotlite` package provides an MIT-licensed UMAP implementation that integrates seamlessly with `embedmit` for complete feature engineering workflows.

## References

- McInnes, L., Healy, J., & Melville, J. (2018). UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction. *arXiv:1802.03426*.
- Kuhn, M., & Silge, J. (2022). *Tidy Modeling with R*. O'Reilly Media.
- [Understanding UMAP](https://pair-code.github.io/understanding-umap/) - Interactive visualization guide
